A Linear Technique to Understand Non-Normal Turbulence Applied to a Magnetized Plasma

### 2014-02-23 Referee_Response1/referee_response.tex

\documentclass[12pt]{article}
\usepackage{graphicx}
\setlength{\parindent}{0in}

% General Latex  --------------------------------------------------
\def\beq{\begin{equation}}
\def\eeq{\end{equation}}
\def\beqar{\begin{eqnarray}}
\def\eeqar{\end{eqnarray}}
\def\nn{\nonumber}
\def\ol{\overline}
\def\para{\parallel}

% Operators  ------------------------------------------------------
\newcommand{\diff}[2]{\frac{d#1}{d#2}}
\newcommand{\diffs}[2]{\frac{d^2#1}{d#2^2}}
\newcommand{\pdiff}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdiffs}[2]{\frac{\partial^2#1}{\partial#2^2}}
\newcommand{\pdiffxy}[3]{\frac{\partial^2#1}{\partial#2 \partial#3}}
\newcommand{\pdt}{\partial_t}
\newcommand{\pdr}{\partial_r}
\newcommand{\pdth}{\partial_\theta}
\newcommand{\pdrr}{\partial^2_r}

\newcommand{\enum}[2]{{#1}\times10^{#2}} % 4.2x10^{3} = \enum{4.2}{3}

\newcommand{\vect}[1]{{\bf #1}}
%\newcommand{\vect}{\overrightarrow}
%\newcommand{\vect}{\vec}
\def\div{\nabla\cdot}
\def\grad{\nabla}
\def\curl{\nabla\times}
\newcommand{\gradpar}{\grad_\parallel}
\newcommand{\gradperp}{\grad_\perp}
\newcommand{\gradr}{\grad_r}
\newcommand{\defeq}{\ensuremath{\stackrel{\text{\tiny def}}{=}}}

\newcommand{\savg}[1]{\left<{#1}\right>}
\newcommand{\vavg}[1]{\left<{#1}\right>_V}
\newcommand{\thavg}[1]{\left<{#1}\right>_\theta}

% Variable names  -------------------------------------------------
\newcommand{\vpar} {v_\parallel}
\newcommand{\Apar} {A_\parallel}
\newcommand{\jpar} {j_\parallel}
\newcommand{\kpar} {k_\parallel}
\newcommand{\kperp} {k_\perp }
\newcommand{\vperp} {v_\perp }
\newcommand{\kthe}{k_\theta}

\newcommand{\Evec}{\ensuremath{\boldsymbol{{\rm E}}}}
\newcommand{\Bvec}{\ensuremath{\boldsymbol{{\rm B}}}}
\newcommand{\Jvec}{\ensuremath{\boldsymbol{{\rm J}}}}
\newcommand{\Fvec}{\ensuremath{\boldsymbol{{\rm F}}}}
\newcommand{\fvec}{\ensuremath{\boldsymbol{{\rm f}}}}
\newcommand{\vE}{\ensuremath{\boldsymbol{{\rm v}_{E}}}}
\newcommand{\bo}{\ensuremath{\boldsymbol{{\rm b}_0}}}
\newcommand{\bvec}{\ensuremath{\boldsymbol{{\rm b}}}}
\newcommand{\xvec}{\ensuremath{\boldsymbol{{\rm x}}}}
\newcommand{\yvec}{\ensuremath{\boldsymbol{{\rm y}}}}
\newcommand{\zvec}{\ensuremath{\boldsymbol{{\rm z}}}}
\newcommand{\vvec}{\ensuremath{\boldsymbol{{\rm v}}}}
\newcommand{\jvec}{\ensuremath{\boldsymbol{{\rm j}}}}

\newcommand{\bxgp}{\bvec\times\gradperp}

\newcommand{\vve}{\ensuremath{\boldsymbol{{\rm v}}_{e}}}
\newcommand{\vvi}{\ensuremath{\boldsymbol{{\rm v}}_{i}}}
\newcommand{\vpe}{v_{\parallel e}}
\newcommand{\vpi}{v_{\parallel i}}
\newcommand{\vvE}{\ensuremath{\boldsymbol{{\rm v}}_{E}}}
\newcommand{\vvD}{\ensuremath{\boldsymbol{{\rm v}}_{D}}}

\newcommand{\nuei}{\nu_{ei}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nue}{\nu_{e}}
\newcommand{\nuen}{\nu_{en}}
\newcommand{\nuin}{\nu_{in}}
\newcommand{\kpe}{\kappa_{\parallel e}}

\newcommand{\rs}{\rho_{s}}
\newcommand{\ri}{\rho_{i}}
\newcommand{\wci}{\Omega_{i}}
\newcommand{\wcix}{\Omega_{ix}}
\newcommand{\wce}{\Omega_{e}}
\newcommand{\tomega}{\tilde\omega}
\newcommand{\Isat}{I_{\rm sat}}
\newcommand{\fmie}{\frac{m_i}{m_e}}
\newcommand{\fmei}{\frac{m_e}{m_i}}

% Often used dimensions
\newcommand{\cm}{\rm cm}
\newcommand{\mm}{\rm mm}
\newcommand{\cmn}{{\rm cm}^{-3}}
\newcommand{\mn}{{\rm m}^{-3}}
\newcommand{\eV}{\rm eV}
\newcommand{\G}{\rm G}
\newcommand{\T}{\rm T}

\begin{document}

{\bf Response to Referees \\
 \\
 \\} 
Brett Friedman \\
Updated: \today \\

\hrulefill

\section{Referee 1}

\emph{1) After advocating the use of non-modal structures rather than linear eigenmodes for the
entirety of the paper, the author use a time-scale derived from linear eigenmodes (see
page 4, paragraph 2). Furthermore, they use the frequency for the n=1 mode instead of
that for the n=0 mode(s), with no apparent justification (other than "nonlinear coupling”).
It would seem much more consistent to use the time-scale for the nonlinear instability
they talk about earlier (although this would mean perhaps a loss of predictability that they
are very keen on). To us, this appears to be the weakest link in the paper. Perhaps, if they
present this paper as a study of the turbulence properties instead, they could give
information on how the curves in Figure 3 change with the choice of the linear time-
scale. It would seem that this would be much more useful to the reader in assessing the
power of the non-modal analysis, rather than presenting just two choices, and claiming
predictability. } \\

Response: \\

\emph{2) What closure assumption is used to write equation (5)?} \\

Response: We do not really understand this question. We do not use any closure to write equation 5. It is an exact equation that can be derived from the model equations~\cite{friedman2012b,friedman2013}. \\

\emph{3) On page 2, last paragraph, there is a puzzling statement: "Perhaps one of the most
useful statistical properties that we can predict is the turbulent growth rate spectrum.” Do
they actually have a measurement of the turbulent growth rate spectrum? Since
predictability is one of their prime concerns, what measurable transport property does the
growth rate spectrum relate to?}

Response: We acknowledge that this was too cryptic a statement. In it's place, we have inserted the following, ``These properties can be calculated through mixing length formulas if the turbulent growth rate spectrum is known.'' Then, as before, we proceed to calculate the turbulent growth rate spectrum from the nonlinear simulation (shown in Figure 1b) and also calculate it from our non-modal linear procedure (labelled $\gamma_{TR}$, the results of which are shown in Figure 3. At the end of the paper, we use the $\gamma_{TR}$ spectrum through the mixing length formula $\gamma/k_\perp^2$ to calculate the turbulent saturation level. \\

\emph{4) We think a log scale should be used for Figure 2. On the scale used in the paper, it is
impossible to determine if the growth rate at short times is somewhat similar to $G_{max} (t)$
or quite different. If similar, the potential for formulating fully analytic models is more
promising.} \\

Response: We have changed Figure 2 to a log scale. The growth rate of the ensemble of curves is not close to the growth rate of $G_{max}(t)$, unfortunately, which is why we did not use it in the calculations of the transient growth rate spectrum. \\

\emph{5) Minor point: Figure 3 is confusing simply because the important curves are not
sufficiently well-delineated. For example, the turbulent growth rate spectrum should be
bold (or something equivalent) to distinguish it from the others.} \\

Response: The curves are different colors, so they should be distinguishable, but not in black and white. Thus, we have put symbols on them. \\

\section{Referee 2}

\emph{1) I find this manuscript somewhat confusing, especially when the authors try to cast it as a ‘non-modal analysis’. For non-modal stability analysis of a linear system,}

\beq
\diff{\mathbf{u}}{t} = \mathbf{A u} \nonumber
\eeq

\emph{I would expect the prediction of the short time behavior of some norm of $\mathbf{u}$ through the mathematical analysis of $\mathbf{A}$
by methods other than the standard eigenmode decomposition, e.g., by methods based on the $\epsilon$-pseudospectrum and/or numerical range.}

\emph{If I understand correctly, the method presented here starts by dropping the nonlinear terms in
the equations of motion of the reduced Braginskii 2-fluid model. The resulting linear system is
numerically simulated up to a time $\tau_{nl}$ at which the system is then ‘re-randomized’ — whatever
it means. A growth rate $\gamma_{TR}$ defined and computed using data between $t = 0$ and $t = \tau_{nl}$ then
constitutes the prediction by this method. So it seems that the ‘non-modal method’ described in
this manuscript is basically a numerical simulation of the linearized equations of motion. While
prediction of turbulent properties by numerical simulations can be useful, it probably has less
impact than a theoretical prediction (which is admittedly much more difficult).} \\

Response: We contend that our method is a form of 'non-modal analysis'. That is, we use only linear calculations to find growth rates without using normal modes (eigenvalues and eigenvectors). While most non-modal analysis in the literature seems to center around the calculation of the $epsilon$-pseudospectrum, much also includes the calculation of $||e^{\mathbf{A} t}||$ and $||e^{\mathbf{A} t} \mathbf{u}(0)||$ with various optimal $\mathbf{u}(0)$ vectors. \\

\emph{2) For the procedure described above, there are several things I do not fully understand. First of
all, if $\gamma_{TR}$ is only computed using data for $0 < t < \tau_{nl}$, how does the ‘re-randomization’ at $t = \tau_{nl}$
come into play? I am probably missing something here.} \\

Response: We agree that this was a confusing way to word our procedure. We therefore have replaced our previous wording with the following: ``we start with an ensemble of random initial conditions, evolve them linearly for a time $\tau_{\rm{nl}}$, and then take the time and ensemble average growth rate of these curves. 
This procedure mimicks what would happen if the turbulence evolved linearly for a time $\tau_{\rm{nl}}$, randomized, and then repeated.''
\\

\emph{3) Secondly, the various growth rates used here are defined as}

\beq
\gamma(m,n) = \frac{Q(m,n) + D(m,n)}{E(m,n)} \nonumber
\eeq
\emph{whereas the evolution of the energy for mode $(m, n)$ is given by}

\beq
\diff{E(m,n)}{t} = Q(m,n) + D(m,n) + \sum_{m',n'} T(m,m',n,n') \nonumber
\eeq
\emph{What is the reason of ignoring $T$ in the definition of $\gamma$? Similarly, what is the meaning/significance
of plotting $E(m, n, t)$ with nonlinearities turned off?  $\sum_{m',n'} T(m,m',n,n')$ is non-zero and contributes to 
$\diff{E(m, n)}{t}$ even though $\sum_{m,m',n,n'} T(m,m',n,n') = 0$. Would it be more informative
to compare $\gamma_{TR}$ to $\diff{E(m, n)}{t}$ instead of the $\gamma_e (m, n)$ defined in the manuscript (which ignores
contribution from $T$ )?} \\

Response: The reason why $T$ is missing from the definition of $\gamma$ is because $\diff{E(m,n)}{t} = Q(m,n) + D(m,n) + \sum_{m',n'} T(m,m',n,n')$ is virtually zero when the turbulence is in steady state (by definition of steady state). It may not be zero at a given time, but it is zero on average, and thus not suitable for use in defining the growth rate spectrum, which is a time-averaged property (see Figure 1b for the average and the spread in the growth rate). If we were to include $T$ in the definition, the turbulent growth rate curves in Figure 1b would be zero with some finite spread. So we can't use $T$ in the definition.

More informatively, the growth rate we use (We didn't invent it. It has been used for many years in numerous papers.) is a measure, by wavenumber, of the rate at which the fluctuations take energy from the equilibrium gradients (or return energy if the growth rate is negative). This comes purely from the linear terms. The nonlinear terms transfer energy between different fluctuation 'modes', but do not affect the overall energy budget.
So the picture is that certain modes or structures (those with positive $\gamma$) inject energy from the equilibrium gradients into the fluctuations through the linear terms. Those modes then non-linearly transfer energy to other modes in an overall conservative way (like a cascade). When energy is nonlinearly transferred into modes with negative growth rate, that energy is dissipated. In the case of normal linear operators, the unstable linear eigenvectors inject energy, transfer that energy to stable linear eigenvectors, which then dissipate the energy. For normal operators, one would expect this growth rate to be pretty similar to the fastest growing eigenmode growth rate, though it could be smaller. But it certainly cannot be larger. We have added and modified quite a bit of text and added a reference~\cite{terry2006b}, which uses this growth rate so that it doesn't appear that we have invented it.

Finally, the reason we plot $E(m,n,t)$ with the nonlinearities turned off is to show that the linear terms alone cause (transient) growth of the fluctuations despite the fact that the $n=0$ mode numbers contain no unstable eigenmodes. It illustrates that the turbulent structures inject energy into the fluctuations in the short term. \\

\emph{4) I also feel that the clarity of the present manuscript suffers because of the page limitation of
PRL, a lot of the symbols are not defined (e.g. $v_E$, $N_r$ ) making it difficult for the general audience.
Together with the concerns raised above, I am afraid I cannot recommend the publication of the
present manuscript in the PRL.} \\

Response: \\

\bibliographystyle{unsrt}
\bibliography{refs}

%
%

\end{document}

### 2014-02-23 Transient_growth_prl.tex
Feb 22

%\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\documentclass[letter,scriptaddress,twocolumn, prl,showkeys]{revtex4}
\usepackage{graphicx}
%\setlength{\parindent}{0in}

@@

\title{A Linear Technique to Predict Non-Normal Turbulence Applied to a Magnetized Plasma}

@@

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative prediction difficult. We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
period consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectrum to that obtained from a nonlinear simulation, showing good quantitative agreement.
\end{abstract}

@@

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are nonorthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under the 
action of the linear operator. Transient growth events, which can dominate nonlinear evolution, require initial-value (non-modal) calculations rather than normal mode calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing a quantitative approach to predicting turbulent properties using only non-modal linear calculations. 
Our approach is to evolve the linear solution from an ensemble of random initial conditions 
and calculate the average growth rate of the solutions over a specified timescale.  Optimally this time scale would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time.
This procedure produces an effective growth rate spectrum that can be used to predict turbulent properties such as saturation levels.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. For this model, the technique reproduces the turbulent growth rate spectrum of the direct nonlinear simulation fairly well, especially
in comparison to the linear eigenmode spectrum.

The model describes pressure-gradient-driven turbulence in the uniformly magnetized, cylindrical plasma
produced by the Large Plasma Device (LAPD)~\cite{gekelman1991}. 
Due to LAPD's large size and high collisionality, a fluid model is
appropriate to describe the turbulence~\cite{Popovich2010a,Popovich2010b,Umansky2011}. We use a reduced Braginskii 2-fluid model:

\beqar
\label{ni_eq}
\pdt N = - {\mathbf v_E} \cdot \grad N_0 - N_0 \gradpar \vpe + S_N + \{\phi,N\}, \\
\label{ve_eq}
\pdt \vpe = - \fmie \frac{T_{e0}}{N_0} \gradpar N - 1.71 \fmie \gradpar T_e  \nonumber \\
+ \fmie \gradpar \phi - \nue \vpe + \{\phi,\vpe \}, \\
\label{rho_eq}
\pdt \varpi = - N_0 \gradpar \vpe  - \nuin \varpi + \{\phi,\varpi \}, \\
\label{te_eq}
\pdt T_e = - {\mathbf v_E} \cdot \grad T_{e0} - 1.71 \frac{2}{3} T_{e0} \gradpar \vpe \nonumber \\
+ \frac{2}{3 N_0} \kpe \gradpar^2 T_e  - \frac{2 m_e}{m_i} \nue T_e  +  S_T + \{\phi,T_e\},
\eeqar
where $N$ is the density, $\vpe$ the parallel electron velocity, $\varpi \equiv \gradperp \cdot (N_0 \gradperp \phi)$ the potential vorticity,
and $T_e$ the electron temperature. All lengths are
normalized to the ion sound gyroradius $\rho_s$, times to the ion cyclotron time $\omega_{ci}^{-1}$, velocities to the sound speed $c_s$, densities to the equilibrium peak density, and electron
temperatures and potentials to the equilibrium peak electron temperature. The profiles $N_0$ and $T_{e0}$ and other parameters are taken from experimental 
measurements~\cite{schaffner2012,friedman2012b,friedman2013}. 
%$\phi_0$  is absent in the equations because we model one particular experiment in which the equilibrium radial electric field was reduced to near zero value using a limiter biasing technique~\cite{schaffner2012}.

The equations are global and partially linearized. The only nonlinearities retained are the advective nonlinearities, which are written with Poisson brackets. 
We add artificial diffusion and viscosity terms with small numerical coefficients ($10^{-3}$)
to ensure numerical stability in nonlinear simulations, which are performed with the BOUT++ code~\cite{dudson2009}. We use periodic axial and zero value radial 
boundary conditions. Further details of the model, including validation studies, may be found in the references~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013},
though we mention here that the model does very well in reproducing
the statistical properties of the experimentally-observed turbulence.

The nonlinear simulation reveals a fascinating property of the turbulence -- 
it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
This nonlinear instability process, which was discovered by Drake et al.~\cite{drake1995} in a similar plasma model, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
These filaments are unstable to a secondary drift waves, which grow on the periphery of the filaments. 
These drift waves, which have finite $k_\para$, nonlinearly couple to one another and reinforce the original convective filaments.

@@

\beq
\label{dEdt_def}
\diff{E(m,n)}{t} = Q(m,n) + D(m,n) + \sum_{m',n'} T(m,m',n,n')
\eeq
where $m$ and $n$ represent the azimuthal and axial Fourier mode numbers. 
%Since Eqs.~\ref{ni_eq}-\ref{te_eq} are not explicitly dependent on $\theta$ and $z$, the energy equation is separable for every $m$ and $n$. 
$Q$ represents the linear energy exchange with the equilibrium (convective transport), $D$ the dissipation from irreversible processes
like collisions, and $T$ the nonlinear triad wave transfer. Since $\int f \{g,f\} dV = 0$, or equivalently $\sum_{m,m',n,n'} T(m,m',n,n')=0$, 
the Poisson bracketed advective nonlinearities in the model are energetically conservative and therefore do no inject or dissipate energy on the whole.
Consequently, the fluctuation energy originates completely from $Q$, the linear convective transport mechanism.

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{n0_n1_growth}}
\caption{a) Linear evolution of energy starting from a turbulent initial state. The $n=0$ curves have an initial period of transient growth before exponentially decaying.
b) Linear and turbulent growth rate spectra for $n=0$ (solid lines) and $n=1$ (dashed lines) Fourier components. The linear growth rates are those of the least stable eigenmodes,
while the turbulent growth rates represent $\pdiff{E}{t}|_{\rm{lin}}/2 E$ from the nonlinear simulation. The shaded region marks the $1 \sigma$ spread in the turbulent spectrum,
obtained from the distribution of growth rates in the nonlinear simulation.}
\label{n0_n1_growth}
\end{figure}

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
streaks~\cite{trefethen1993,krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative modes are shown in Fig.~\ref{n0_n1_growth} a). The linear transient growth of the filamentary $n=0$ structures is evident as their modes 
grow transiently before decaying exponentially at the rate of their least stable eigenmode. Such behavior is rather unituitive because all $n=0$ linear eigenmodes are stable, 
and the normal mode paradigm has conditioned us to believe that a linear superposition of such eigenmodes should decay. But as the hydrodynamics community discovered, linear
superpositions of nonorthogonal vectors may behave unintuitively.
Notice finally that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode.
%This isn't necessarily due to eigenvector nonorthogonality, but is nevertheless unintuitive.

>>>>

%\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\documentclass[letter,scriptaddress,twocolumn, prl,showkeys]{revtex4}
\usepackage{graphicx}
%\usepackage{amssymb}
%\setlength{\parindent}{0in}

@@

\title{A Linear Technique to Understand Non-Normal Turbulence Applied to a Magnetized Plasma}

@@

\begin{abstract}
In nonlinear dynamical systems with highly nonorthogonal linear eigenvectors, linear non-modal analysis is more useful than normal mode analysis in predicting turbulent properties. 
However, the non-trivial time evolution of non-modal structures makes quantitative prediction difficult. We present a technique to overcome this difficulty by modelling the effect that the advective nonlinearities have on spatial turbulent structures. The nonlinearities are taken as a periodic randomizing force with
period consistent with critical balance arguments. We apply this technique to a model of drift wave turbulence in the Large Plasma Device (LAPD) 
[W. Gekelman \emph{et al.}, Rev. Sci. Inst. {\bf 62}, 2875 (1991)], where non-modal effects dominate the turbulence.
We compare the resulting growth rate spectra to that obtained from a nonlinear simulation, showing good qualitative agreement.
\end{abstract}

@@

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are nonorthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under the 
action of the linear operator. Transient growth behavior, which can dominate turbulent evolution, can be discovered only through non-modal calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
still heavily relies on normal mode analysis to inform turbulent predictions and observations, with a few notable exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions. 
Our approach is to evolve the linear solution from an ensemble of random initial conditions 
and calculate the average growth rate of the solutions over a specified timescale.  Optimally this time scale would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since several linear time scales exist, we test them all and compare the results to determine which works best.
The procedure ultimately produces an effective growth rate spectrum that can be used to predict turbulent properties such as saturation levels through mixing length arguments.
While the concepts behind this technique are general enough to be applied to many nonlinear dynamical systems, the details vary for each case, 
so we restrict our treatment to one particular turbulence model. For this model, the technique qualitatively reproduces the turbulent growth rate spectrum of the direct 
nonlinear simulation fairly well, especially in comparison to the linear eigenmode spectrum.

The model describes pressure-gradient-driven turbulence in the uniformly magnetized, cylindrical plasma
produced by the Large Plasma Device (LAPD)~\cite{gekelman1991}. 
We use a reduced Braginskii 2-fluid model~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013}:

\beqar
\label{ni_eq}
\pdt N = - {\mathbf v_E} \cdot \grad N_0 - N_0 \gradpar \vpe + S_N + \{\phi,N\}, \\
\label{ve_eq}
\pdt \vpe = - \fmie \frac{T_{e0}}{N_0} \gradpar N - 1.71 \fmie \gradpar T_e  \nonumber \\
+ \fmie \gradpar \phi - \nue \vpe + \{\phi,\vpe \}, \\
\label{rho_eq}
\pdt \varpi = - N_0 \gradpar \vpe  - \nuin \varpi + \{\phi,\varpi \}, \\
\label{te_eq}
\pdt T_e = - {\mathbf v_E} \cdot \grad T_{e0} - 1.71 \frac{2}{3} T_{e0} \gradpar \vpe \nonumber \\
+ \frac{2}{3 N_0} \kpe \gradpar^2 T_e  - \frac{2 m_e}{m_i} \nue T_e  +  S_T + \{\phi,T_e\},
\eeqar
where $N$ is the density, $\vpe$ the parallel electron velocity, $\varpi \equiv \gradperp \cdot (N_0 \gradperp \phi)$ the potential vorticity,
$T_e$ the electron temperature, ${\mathbf v_E}$ the ${\mathbf E} \times {\mathbf B}$ velocity, and $S_N$ and $S_T$ density and temperature sources. All lengths are
normalized to the ion sound gyroradius $\rho_s$, times to the ion cyclotron time $\omega_{ci}^{-1}$, velocities to the sound speed $c_s$, 
densities to the equilibrium peak density, and electron temperatures and potentials to the equilibrium peak electron temperature. 
The profiles $N_0$ and $T_{e0}$ and other parameters are taken from experimental measurements~\cite{schaffner2012,friedman2012b,friedman2013}.

The equations are global and we retain only advective nonlinearities, which are written with Poisson brackets. 
We add artificial diffusion and viscosity terms with small numerical coefficients ($10^{-3}$)
to ensure numerical stability in nonlinear simulations, which are performed with the BOUT++ code~\cite{dudson2009}. We use periodic axial and zero value radial 
boundary conditions. Further details of the model, including validation studies, may be found in the references~\cite{Popovich2010a,Popovich2010b,Umansky2011,friedman2012b,friedman2013},
though we mention here that the model does very well in reproducing the statistical properties of the experimentally-observed turbulence.

The nonlinear simulation reveals a fascinating property of the turbulence -- 
it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
The nonlinear instability, which was discovered by Drake et al.~\cite{drake1995}, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
These filaments are unstable to a secondary drift waves, which grow on the periphery of the filaments. 
These drift waves, which have finite $k_\para$, nonlinearly couple to one another and reinforce the original convective filaments.

@@

\beq
\label{dEdt_def}
\diff{E(m,n)}{t} = \diff{E_{l}(m,n)}{t} + \diff{E_{nl}(m,n)}{t}
\eeq
where $m$ and $n$ represent the azimuthal and axial Fourier mode numbers. 
$\diff{E_{l}(m,n)}{t}$ comes from the linear terms in Eqs.~\ref{ni_eq}-\ref{te_eq}, $\diff{E_{nl}(m,n)}{t}$ comes from the nonlinear terms. 
$\diff{E_{l}(m,n)}{t}$ describes the injection (or dissipation) of energy into the fluctuations from the free energy in the equilibrium gradients.
$\diff{E_{nl}(m,n)}{t}$ describes the energy exchange between fluctuations with different $m,n$ and it is conservative: $\sum_{m,n} \diff{E_{nl}(m,n)}{t} = 0$.
Moreover, in quasi-steady state turbulence, the rate of energy injection (or dissipation) into the fluctuations at each $m,n$ by the linear terms must be balanced by
the rate of energy removal (or deposition) from the nonlinear terms:

\beqar
\label{steady_state}
\gamma(m,n) \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE(m,n)/dt}{2 E(m,n)} dt \nonumber \\
= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E(T)/E(0) \right]}{T} = 0 \quad {\rm{in \ steady \ state}}.
\eeqar

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{n0_n1_growth}}
\caption{a) Linear evolution of energy starting from a turbulent initial state. The $n=0$ curves have an initial period of transient growth before exponentially decaying.
b) Linear and turbulent growth rate spectra for $n=0$ (solid lines) and $n=1$ (dashed lines) Fourier components. The linear growth rates are those of the least stable eigenmodes,
while the turbulent growth rates represent $\pdiff{E_l}{t}/2 E$ from the nonlinear simulation. The shaded region marks the $1 \sigma$ spread in the turbulent spectrum,
obtained from the distribution of growth rates in the nonlinear simulation.}
\label{n0_n1_growth}
\end{figure}

From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma(m,n) = \gamma_l(m,n) + \gamma_{nl}(m,n) = 0$.
The rate of energy injection $\gamma_l(m,n)$ from the equilibrium gradient into the fluctuations is a quantity of great interest~\cite{friedman2012b,terry2006b}. 
When positive for some wavenumber, turbulence can be sustained. Additionally, it may be used in the mixing length formula $\gamma/k_\perp^2$ to predict the turbulent saturation level, 
and it is proportional to the transport rate~\cite{terry2006b}. 
$\gamma_l(m,n)$ may be calculated from the spatial structures of of the plasma state variables, so it has meaning in turbulent plasma states. We plot
$\gamma_l(m,n)$ in Fig.~\ref{n0_n1_growth} a) for both the linear and steady state turbulent stages of our nonlinear simulation. The linear stage, which occurs when fluctuations
are small and exponentially growing has a time-independent $\gamma_l(m,n)$ equal to $\gamma_s(m,n)$ -- the ``spectral'' growth rate of the fastest growing eigenmode at each $m,n$.
During the turbulent stage, $\gamma_l(m,n)$ is time-dependent (indicated by the $1 \sigma$ spread), and generally much different than $\gamma_s(m,n)$. Significantly,
the turbulent $\gamma_l$ is positive at $n=0$ and low $m$ despite the fact that all linear eigenmodes have $\gamma_s < 0$ for $n=0$. This is a manifestation
of non-normality, for in normal systems, $\gamma_l(m,n) \le \gamma_s(m,n)$. Physically, it is the manifestation of the nonlinear
instability, specifically the linear part of the mechanism in which convective filaments drive density filaments from the equilibrium density gradient.

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
streaks~\cite{trefethen1993,krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative modes are shown in Fig.~\ref{n0_n1_growth} b). The linear transient growth of the filamentary $n=0$ structures is evident as their modes 
grow transiently before decaying exponentially at the rate of their least stable eigenmode. Such behavior is indicative of non-modal behavior. It must be since 
all $n=0$ linear eigenmodes are stable. Notice also that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode. This transient
is another non-modal result.

Since transient growth is a purely linear phenomenon, it has been a goal of researchers to understand and predict the onset of subcritical turbulence using only linear, non-modal calculations. 
In our system, the turbulence is not subcritical in the traditional sense, but it has a subcritical component in the $n=0$ modes. They have no unstable eigenmodes, but still inject energy into the
turbulence -- $\gamma_l(m<50,n=0) > 0$, yet $\gamma_s(m,n=0) < 0$. It is our goal, then, to use linear non-modal calculations to understand this behavior and to move toward predictive capability
of $\gamma_l$.
To accomplish this, we make the anzatz that nonlinearities randomize the turbulent spatial structure at each wavenumber on a time scale of 
one eddy decorrelation time, while the linearities evolve the spatial structures deterministically. From this we can calculate a ``Transient Growth Rate'' spectrum,
which is our prediction of the turbulent growth rate spectrum. To illustrate,
we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.  Then, we discritize in the radial direction and
approximate radial derivatives with finite differences.  The resulting system of equations may be written in matrix form:
